A fast GPU solver for near-Toeplitz tridiagonal systems

### 2015-11-08 paper-draft/paper.tex
change Figure to Fig.

Cyclic reduction is applicable to the solution of
general linear systems of the form $A\bm{x} = \bm{d}$,
where $A$ is a tridiagonal matrix with diagonals
$\bm{a}$, $\bm{b}$ and $\bm{c}$
(Equation \ref{eqn:general-tridiagonal-system}).
The algorithm consists of two phases:
\emph{forward reduction} and \emph{backward substitution}
\ref{fig:cyclic-reduction}.
In the forward reduction phase,
every even-indexed equation $i$
is expressed as a
linear combination of equations $i$, $i-1$ and $i+1$,
yielding a new tridiagonal system of
$n/2$ equations in $n/2$ unknowns
(Equations \ref{eqn:forward-reduction-1} - \ref{eqn:forward-reduction-4}).
The process is repeated until a system of
2 equations in 2 unknowns is left.

@@

As described in Section \ref{subsec:gpu-architecture},
GPU instructions are executed concurrently
by several threads and
threads are organized into thread blocks.
In most GPU implementations of cyclic reduction,
blocks are assigned to tridiagonal systems
(when solving multiple systems),
and threads within a block 
are assigned to equations, or \emph{indices}.
(Figure \ref{fig:gpu-mapping})
During the forward reduction phase,
the threads assigned to each even index $i$
compute the coefficients
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$.
In practice, the coefficients are updated \emph{in-place}
(Figure \ref{fig:forward-reduction-step} shows the updates
to the coefficient array $\bm{b}$ in the first forward reduction step).

@@

The method is based on precomputing elements 
appearing in the forward reduction phase
so that they can be reused for multiple right hand sides.
While this can be done for general tridiagonal systems,
the applicability to near-Toeplitz systems is
of particular interest due to the lower storage costs.
In Figure \ref{fig:forward-reduction-step},
we list the updates to
$b_i$, for $i=2, 4, 6 ... n$
during the first forward reduction step
for a general tridiagonal matrix.
Similar equations can be written describing
the updates to $a_i$ and  $c_i$.

@@

At each reduction step,
for every index $i$,
the coefficients $a_i^\prime$, $b_i^\prime$, $c_i^\prime$
and auxiliary variables $k_1$ and $k_2$
(hereby denoted as $k_{1,i}$ and $k_{2,i}$) are identical,
except for those influenced by the
boundary coefficients of the previous step.
It is possible to \emph{precompute} these values
and store them in separate arrays.
In Figure \ref{fig:cyclic-reduction-precomputing},
it is seen that the space required for storing each of the
forward reduction coefficients
$a_i^\prime$, $b_i^\prime$, $c_i^\prime$,
$k_{1,i}$ and $k_{2,i}$
for a near-Toeplitz matrix
is, at most, $log_2n-2 + 2(log_2n-1)$.

@@

Following the algorithm described in \ref{sec:proposed-algorithm},
the forward reduction coefficients
are first computed and loaded into global memory.
The right hand sides are stored in a single
contiguous region global memory.
In both approaches,
we map each of the $N_{rhs}$ tridiagonal systems
to a thread block,
and equations or \emph{indices} of a system
to individual threads within a block
(Figure \ref{fig:gpu-mapping}).

@@

In the shared memory approach (Figure \ref{fig:global-and-shared}),
we launch a single kernel to perform the entire
cyclic reduction solve.
The kernel is launched with $n/2$ threads per block.
Each thread block allocated a block of shared memory of size $n/2$.
Each thread performs the first reduction step
(Equation \ref{eqn:precomputed-forward-reduction-step})
by accessing the required values
$d_i$, $d_{i-1}$ and $d_{i+1}$ from global memory,
storing the result in shared memory.
In subsequent reduction steps,
$d_i$, $d_{i-1}$ and $d_{i+1}$
are accessed from shared memory,
avoiding the uncoalesced global memory accesses
seen in the global memory implementation.
In each back substitution step,
threads overwrite the existing values in shared memory
with the values of the solution.
In the final step,
shared memory is filled completeley
with the even-indexed solution values.
Each thread then computes an odd-indexed solution value,
storing it directly in global memory
and copies the even-indexed solution value
from shared memory to global memory.
The entire solution is done in a single kernel launch
to hide the latency of
memory transfers between global and shared memory.
Explicit synchronization between the threads of a block
is required within the kernel at the end of each step.

@@

In Figures \ref{fig:global-vs-shared-2d} and \ref{fig:global-vs-shared-3d},
we report the performance of the two solvers for the case
$N_{rhs} = n$ and $N_{rhs} = n^2$.
These cases correspond to tridiagonal systems
arising in 2-D and 3-D problems respectively.
We note that the shared memory implementation
offers better performance in nearly all cases.
However, the relative speedup from using
shared memory diminishes with increasing problem size.
For larger problem sizes,
the reduced occupancy and increase in number of inactive threads
leads to poor shared memory performance.

@@

In Figures \ref{fig:bench-2d} and \ref{fig:bench-3d},
we provide the relative performance of
Intel MKL and CUSPARSE solvers and compare against
the NEATO shared memory implementation.
The Intel MKL solver can use multiple threads
to solve several systems simultaneously.
The CUSPARSE solver uses a hybrid algorithm composed
of Cyclic Reduction and Parallel Cyclic Reduction (CR+PCR).
The relative performance for
each problem size is obtained by
normalizing the solver timings
by the timing for the fastest solver for that problem size.
Table \ref{table:bench} shows the timings of the various
solvers to solve different problem sizes.

@@

The computational domain (Figure \ref{fig:computational-domain})
is a structured grid comprised of
$nx \times ny \times nz$ grid points.
While evaluating the derivative in, say, the x-direction,
the computational domain is thought of as
a collection of \emph{grid lines}---each consisting of
$nx$ grid points---oriented in the x-direction.
For each grid line,
we may assemble a tridiagonal system of size $nx$, as in
Equation \ref{eqn:compact-tridiagonal-system}.
Solving for the all derivatives in the $x$ direction then involves
solving this tridiagonal system for $ny \times nz$ right hand sides.

@@

The set up and solution of the reduced system
requires global communication of the boundary information,
which may be associated with a high overhead.
While several techniques have been proposed to implement
this type of communication pattern efficiently,
we find it sufficient for moderately large problems
to communicate all information to rank 0,
using \texttt{MPI\_Gather}, and distribute the results
back using \texttt{MPI\_Scatter} (Figure \ref{fig:reduced-system}).

>>>>

Cyclic reduction is applicable to the solution of
general linear systems of the form $A\bm{x} = \bm{d}$,
where $A$ is a tridiagonal matrix with diagonals
$\bm{a}$, $\bm{b}$ and $\bm{c}$
(Equation \ref{eqn:general-tridiagonal-system}).
The algorithm consists of two phases:
\emph{forward reduction} and \emph{backward substitution}
(Fig. \ref{fig:cyclic-reduction}).
In the forward reduction phase,
every even-indexed equation $i$
is expressed as a
linear combination of equations $i$, $i-1$ and $i+1$,
yielding a new tridiagonal system of
$n/2$ equations in $n/2$ unknowns
(Equations \ref{eqn:forward-reduction-1} - \ref{eqn:forward-reduction-4}).
The process is repeated until a system of
2 equations in 2 unknowns is left.

@@

As described in Section \ref{subsec:gpu-architecture},
GPU instructions are executed concurrently
by several threads and
threads are organized into thread blocks.
In most GPU implementations of cyclic reduction,
blocks are assigned to tridiagonal systems
(when solving multiple systems),
and threads within a block 
are assigned to equations, or \emph{indices}
(Fig. \ref{fig:gpu-mapping}).
During the forward reduction phase,
the threads assigned to each even index $i$
compute the coefficients
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$.
In practice, the coefficients are updated \emph{in-place}.
Figure \ref{fig:forward-reduction-step} shows the updates
to the coefficient array $\bm{b}$ in the first forward reduction step).

@@

The method is based on precomputing elements 
appearing in the forward reduction phase
so that they can be reused for multiple right hand sides.
While this can be done for general tridiagonal systems,
the applicability to near-Toeplitz systems is
of particular interest due to the lower storage costs.
In Fig. \ref{fig:forward-reduction-step},
we list the updates to
$b_i$, for $i=2, 4, 6 ... n$
during the first forward reduction step
for a general tridiagonal matrix.
Similar equations can be written describing
the updates to $a_i$ and  $c_i$.

@@

At each reduction step,
for every index $i$,
the coefficients $a_i^\prime$, $b_i^\prime$, $c_i^\prime$
and auxiliary variables $k_1$ and $k_2$
(hereby denoted as $k_{1,i}$ and $k_{2,i}$) are identical,
except for those influenced by the
boundary coefficients of the previous step.
It is possible to \emph{precompute} these values
and store them in separate arrays.
In Fig. \ref{fig:cyclic-reduction-precomputing},
it is seen that the space required for storing each of the
forward reduction coefficients
$a_i^\prime$, $b_i^\prime$, $c_i^\prime$,
$k_{1,i}$ and $k_{2,i}$
for a near-Toeplitz matrix
is, at most, $log_2n-2 + 2(log_2n-1)$.

@@

Following the algorithm described in \ref{sec:proposed-algorithm},
the forward reduction coefficients
are first computed and loaded into global memory.
The right hand sides are stored in a single
contiguous region global memory.
In both approaches,
we map each of the $N_{rhs}$ tridiagonal systems
to a thread block,
and equations or \emph{indices} of a system
to individual threads within a block
(Fig. \ref{fig:gpu-mapping}).

@@

In the shared memory approach (Fig. \ref{fig:global-and-shared}),
we launch a single kernel to perform the entire
cyclic reduction solve.
The kernel is launched with $n/2$ threads per block.
Each thread block allocated a block of shared memory of size $n/2$.
Each thread performs the first reduction step
(Equation \ref{eqn:precomputed-forward-reduction-step})
by accessing the required values
$d_i$, $d_{i-1}$ and $d_{i+1}$ from global memory,
storing the result in shared memory.
In subsequent reduction steps,
$d_i$, $d_{i-1}$ and $d_{i+1}$
are accessed from shared memory,
avoiding the uncoalesced global memory accesses
seen in the global memory implementation.
In each back substitution step,
threads overwrite the existing values in shared memory
with the values of the solution.
In the final step,
shared memory is filled completeley
with the even-indexed solution values.
Each thread then computes an odd-indexed solution value,
storing it directly in global memory
and copies the even-indexed solution value
from shared memory to global memory.
The entire solution is done in a single kernel launch
to hide the latency of
memory transfers between global and shared memory.
Explicit synchronization between the threads of a block
is required within the kernel at the end of each step.

@@

In Figs. \ref{fig:global-vs-shared-2d} and \ref{fig:global-vs-shared-3d},
we report the performance of the two solvers for the case
$N_{rhs} = n$ and $N_{rhs} = n^2$.
These cases correspond to tridiagonal systems
arising in 2-D and 3-D problems respectively.
We note that the shared memory implementation
offers better performance in nearly all cases.
However, the relative speedup from using
shared memory diminishes with increasing problem size.
For larger problem sizes,
the reduced occupancy and increase in number of inactive threads
leads to poor shared memory performance.

@@

In Fig. \ref{fig:bench-2d} and \ref{fig:bench-3d},
we provide the relative performance of
Intel MKL and CUSPARSE solvers and compare against
the NEATO shared memory implementation.
The Intel MKL solver can use multiple threads
to solve several systems simultaneously.
The CUSPARSE solver uses a hybrid algorithm composed
of Cyclic Reduction and Parallel Cyclic Reduction (CR+PCR).
The relative performance for
each problem size is obtained by
normalizing the solver timings
by the timing for the fastest solver for that problem size.
Table \ref{table:bench} shows the timings of the various
solvers to solve different problem sizes.

@@

The computational domain (Fig. \ref{fig:computational-domain})
is a structured grid comprised of
$nx \times ny \times nz$ grid points.
While evaluating the derivative in, say, the x-direction,
the computational domain is thought of as
a collection of \emph{grid lines}---each consisting of
$nx$ grid points---oriented in the x-direction.
For each grid line,
we may assemble a tridiagonal system of size $nx$, as in
Equation \ref{eqn:compact-tridiagonal-system}.
Solving for the all derivatives in the $x$ direction then involves
solving this tridiagonal system for $ny \times nz$ right hand sides.

@@

The set up and solution of the reduced system
requires global communication of the boundary information,
which may be associated with a high overhead.
While several techniques have been proposed to implement
this type of communication pattern efficiently,
we find it sufficient for moderately large problems
to communicate all information to rank 0,
using \texttt{MPI\_Gather}, and distribute the results
back using \texttt{MPI\_Scatter} (Fig. \ref{fig:reduced-system}).
